---
title: "Liquor Sales VS DUI Accidents"
author: "Gaurav Madarkal"
date: "8/15/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Liquor Sales Vs DUI Accidents

## Installing required dependencies

```{r}
library(dplyr)
library(usmap)
library(ggplot2)
```

## Loading Data - Accident reports - Year wise

```{r}

# INDIVIDUAL YEAR DATA

accidents_2014 <- read.csv('./Datasets/accident_2014.csv')[ ,c('STATE', 'YEAR', 'DRUNK_DR')]
accidents_2015 <- read.csv('./Datasets/accident_2015.csv')[ ,c('STATE', 'YEAR', 'DRUNK_DR')]
accidents_2016 <- read.csv('./Datasets/accident_2016.csv')[ ,c('STATE', 'YEAR', 'DRUNK_DR')]
accidents_2017 <- read.csv('./Datasets/accident_2017.csv')[ ,c('STATE', 'YEAR', 'DRUNK_DR')]
accidents_2018 <- read.csv('./Datasets/accident_2018.csv')[ ,c('STATE', 'YEAR', 'DRUNK_DR')]
#accidents_2019 <- read.csv('./Datasets/accident_2019.csv')[ ,c('STATE', 'YEAR', 'DRUNK_DR')]

# STATE NUMBERS
state_numbers_data <- read.csv('./Datasets/state_names_numbers.csv')

```

## Loading Data - Census - Year wise

```{r}

# GET THE POPULATION DATA FROM YEAR 2015 TO 2019 SO THAT WE CAN CALCULATE THE GALLONS ON ALCOHOL CONSUMED/SOLD STATE WISE
population_state_year <- read.csv('./Datasets/population_statewise_2014-2018.csv', check.names = FALSE)[, c("Name", "2014", "2015", '2016', "2017", "2018")]

# GET THE ALCOHOL CONSUMED DATA, WHICH HAS ETHANOL CONSUMED PER CAPITA
liquor_sales_raw <- read.csv('./Datasets/apparent_per_capita_liquor_sales_1977_2018.csv')

```

## Filtering Data and cleaning it

```{r}

# COMBINE ALL REPORTS FROM EACH YEAR
accidents_all <- rbind(accidents_2014, accidents_2015, accidents_2016, accidents_2017, accidents_2018)

# FILTER ACCIDENTS WHICH ARE DRUNK DRIVE ONLY
dui_accidents_all <- subset(accidents_all, DRUNK_DR == 1)

# RETRIEVE THE STATENAMES BY JOINING WITH STATE NUMBERS
accidents_state_wise <- merge(dui_accidents_all, state_numbers_data, all.x = FALSE, all.y = FALSE)
accidents_state_wise$region <- tolower(accidents_state_wise$STATE_NAME)

# FILTER THE REQUIRED YEARS AND COLUMNS
liquor_sales_2014_2018 <- subset(liquor_sales_raw, year %in% c("2014", "2015", "2016", "2017", "2018"))[, c('state', 'year', 'ethanol_all_drinks_gallons_per_capita')]
```

## Calculating Overall Alcohol Sales per state

```{r}
# NORMALIZE THE STATE NAMES BY MAKING THEM UPPER CASE
liquor_sales_2014_2018$region <- tolower(liquor_sales_2014_2018$state)

# NORMALIZE THE STATE NAMES BY MAKING THEM UPPER CASE
population_state_year$region <- tolower(population_state_year$Name)

# DECLARE ALL THE STATE POPULATION AS 0 INITIALLY
liquor_sales_2014_2018[, "population"] = 0

# LOOP OVER THE INDIVIDUAL STATE POPULATION AND ADD THE VALUES TO ALCOHOL SALES DATA SET
# SINCE THE POPULATION DATASET IS IN A MATRIX FORMAT WE NEED TO LOOP OVER EACH ENTRY AND CANNOT AUTOMATICALLY JOIN
i <- 1
for (row in 1:nrow(population_state_year)) {
  if (i < nrow(liquor_sales_2014_2018)) {
      liquor_sales_2014_2018[i, "population"] = population_state_year[row, "2018"]
      liquor_sales_2014_2018[i + 1, "population"] = population_state_year[row, "2017"]
      liquor_sales_2014_2018[i + 2, "population"] = population_state_year[row, "2016"]
      liquor_sales_2014_2018[i + 3, "population"] = population_state_year[row, "2015"]
      liquor_sales_2014_2018[i + 4, "population"] = population_state_year[row, "2014"]
      i <- i + 5
  }
}
# REMOVE COMMAS AND CONVERT THE POPULATION VALUE TO A NUMERIC ONE
liquor_sales_2014_2018$population <-  as.numeric(gsub(",","",liquor_sales_2014_2018$population))

# MULIPLY THE POPULATION WITH THE PER CAPITA sales WHICH WILL GIVE US THE TOTAL GALLONS A STATE CONSUMED IN THAT YEAR
liquor_sales_2014_2018$total_gallons <- liquor_sales_2014_2018$ethanol_all_drinks_gallons_per_capita * liquor_sales_2014_2018$population
```

## Aggregating total DUI accidents for a year for a state

```{r}

accidents_year_vs_state <- setNames(aggregate(accidents_state_wise$DRUNK_DR, by=list(state=accidents_state_wise$STATE_NAME, year=accidents_state_wise$YEAR), FUN=sum), c("state", "year", "occurences"))
accidents_year_vs_state$fips = fips(accidents_year_vs_state$state)
map_data <- inner_join(us_map(), accidents_year_vs_state)

### ACCIDENTS USA HEATMAP FOR 2014 ###
plot_usmap(data = subset(map_data, year=='2014') %>% select(fips, occurences), values="occurences", labels = TRUE) + 
    scale_fill_continuous(low = "white", high = "red", name = "DUI Accidents 2014", label = scales::comma)
ggsave('./Plots/dui_accidents_2014.png', last_plot())

### ACCIDENTS USA HEATMAP FOR 2015 ###
plot_usmap(data = subset(map_data, year=='2015') %>% select(fips, occurences), values="occurences", labels = TRUE) + 
    scale_fill_continuous(low = "white", high = "red", name = "DUI Accidents 2015", label = scales::comma)
ggsave('./Plots/dui_accidents_2015.png', last_plot())

### ACCIDENTS USA HEATMAP FOR 2016 ###
plot_usmap(data = subset(map_data, year=='2016') %>% select(fips, occurences), values="occurences", labels = TRUE) + 
    scale_fill_continuous(low = "white", high = "red", name = "DUI Accidents 2016", label = scales::comma)
ggsave('./Plots/dui_accidents_2016.png', last_plot())

### ACCIDENTS USA HEATMAP FOR 2017 ###
plot_usmap(data = subset(map_data, year=='2017') %>% select(fips, occurences), values="occurences", labels = TRUE) + 
    scale_fill_continuous(low = "white", high = "red", name = "DUI Accidents 2017", label = scales::comma)
ggsave('./Plots/dui_accidents_2017.png', last_plot())

### ACCIDENTS USA HEATMAP FOR 2018 ###
plot_usmap(data = subset(map_data, year=='2018') %>% select(fips, occurences), values="occurences", labels = TRUE) + 
    scale_fill_continuous(low = "white", high = "red", name = "DUI Accidents 2018", label = scales::comma)
ggsave('./Plots/dui_accidents_2018.png', last_plot())
```

## Plotting USA heat map to understand which states have higher and lower amount of alcohol sales

```{r}

liquor_sales_2014_2018$fips = fips(liquor_sales_2014_2018$state)
map_data <- inner_join(us_map(), liquor_sales_2014_2018)

### ALCOHOL sales FOR YEAR 2014 ###
plot_usmap(data = subset(map_data, year == "2014") %>% select(fips, total_gallons), values="total_gallons", labels = TRUE) + 
    scale_fill_continuous(low = "white", high = "blue", name = "Alcohol sales 2014", label = scales::comma)
ggsave('./Plots/liqour_sales_2014.png', last_plot())


### ALCOHOL sales FOR YEAR 2015 ###
plot_usmap(data = subset(map_data, year == "2015") %>% select(fips, total_gallons), values="total_gallons", labels = TRUE) + 
    scale_fill_continuous(low = "white", high = "blue", name = "Alcohol sales 2015", label = scales::comma)
ggsave('./Plots/liqour_sales_2015.png', last_plot())


### ALCOHOL sales FOR YEAR 2016 ###
plot_usmap(data = subset(map_data, year == "2016") %>% select(fips, total_gallons), values="total_gallons", labels = TRUE) + 
    scale_fill_continuous(low = "white", high = "blue", name = "Alcohol sales 2016", label = scales::comma)
ggsave('./Plots/liqour_sales_2016.png', last_plot())


### ALCOHOL sales FOR YEAR 2017 ###
plot_usmap(data = subset(map_data, year == "2017") %>% select(fips, total_gallons), values="total_gallons", labels = TRUE) + 
    scale_fill_continuous(low = "white", high = "blue", name = "Alcohol sales 2017", label = scales::comma)
ggsave('./Plots/liqour_sales_2017.png', last_plot())


### ALCOHOL sales FOR YEAR 2018 ###
plot_usmap(data = subset(map_data, year == "2018") %>% select(fips, total_gallons), values="total_gallons", labels = TRUE) + 
    scale_fill_continuous(low = "white", high = "blue", name = "Alcohol sales 2018", label = scales::comma)
ggsave('./Plots/liqour_sales_2018.png', last_plot())

```
